---
title: "sentiment.ai"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sentiment.ai}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sentiment.ai)
```

`sentiment.ai` turns text into sentiment scores using a sentence-embedding model
plus a small, bundled scoring head. The default backend is the on-device
multilingual model `multilingual-e5-small` — **no TensorFlow and no API key**.

**Project page:**
[https://benwiseman.github.io/sentiment.ai/](https://benwiseman.github.io/sentiment.ai/)

# Quick start

```r
# one-time setup of the (TensorFlow-free) Python backend
install_sentiment.ai()

# load the default model, then score some text
init_sentiment.ai(model = "e5-small")
sentiment_score(c("I love this!", "this is terrible"))
#> [1]  0.9  -0.9
```

Scores are rescaled to `[-1, 1]` — about `1` is positive, about `-1` is negative.

# Why embeddings?

Compared with lexicon/dictionary sentiment methods, this approach:

1. **Is more robust** — tolerates spelling mistakes and mixed case, and the default
   model handles ~100 languages.
2. **Doesn't need a rigid lexicon** — text is reduced to an embedding vector, so you
   get sensible scores for words and phrasings no dictionary lists.
3. **Lets you choose the context** — `sentiment_match()` lets you define what
   *positive* and *negative* mean for your domain.
4. **Is auditable** — scoring is deterministic, and `sentiment_provenance()` reports
   exactly which model and head produced a score.

# Installation & setup

After installing from CRAN, set up the Python backend once with
`install_sentiment.ai()`. The default install is TensorFlow-free. Useful arguments:

* `envname` — name of the Python environment (default `"r-sentiment-ai"`).
* `method` — `"auto"` (default), `"virtualenv"`, or `"conda"`.
* `gpu` — `FALSE` (default) installs the smaller **CPU-only** PyTorch; `TRUE` installs the
  CUDA build (Linux/Windows with an NVIDIA GPU). macOS is always CPU/MPS. The setup
  walkthrough offers this as a menu choice.
* `fast` — use `uv` (if installed) for a much faster install; falls back to pip safely.
* `legacy` — set `TRUE` to also install TensorFlow / TF-Hub for the opt-in legacy
  Universal Sentence Encoder models.

```r
install_sentiment.ai()                 # TensorFlow-free default
# install_sentiment.ai(legacy = TRUE)  # opt into the legacy USE models
```

Even simpler for newcomers: you can skip this step entirely. The first time you score
text in an interactive session, sentiment.ai offers a short setup walkthrough
(`setup_sentiment.ai()`) and — when [uv](https://docs.astral.sh/uv/) is installed — uses it
for a much faster install. In scripts / non-interactive sessions nothing is auto-installed;
you get a clear message to run `install_sentiment.ai()`.

Once the backend is set up you don't need to call `init_sentiment.ai()` yourself — the
model loads automatically on the first `sentiment_score()` / `sentiment()` /
`sentiment_match()` call. To load it eagerly the moment you attach the package (e.g. from
your `.Rprofile`), set `options(sentiment.ai.autoinit = TRUE)`.

# Choosing a model

`init_sentiment.ai(model = ...)` loads a model into memory:

* `"e5-small"` — default, 384-d, multilingual, on-device, no TensorFlow.
* `"e5-base"` — 768-d, best on-device quality.
* `"openai"` — `text-embedding-3-small` (paid API; text leaves your machine).
* `"en"` / `"en.large"` / `"multi"` / `"multi.large"` — legacy USE models, require
  `install_sentiment.ai(legacy = TRUE)`.

```r
init_sentiment.ai()                    # e5-small
init_sentiment.ai(model = "e5-base")   # larger on-device model
```

# Sentiment analysis

## `sentiment_score()`

The main function — one score per input. The scoring head is `"mlp"` (default) or
`"logistic"`; both are small pure-R JSON heads (no xgboost, no TensorFlow at score
time).

```r
sentiment_score(c("Will you marry me?", "Oh, you're breaking up with me..."))
```

## `sentiment()`

When you need the whole picture rather than one number — the per-class probabilities, a
label, and a confidence — use `sentiment()`. It returns a tidy data frame.

```r
sentiment(c("I love this!", "it's fine", "this is terrible"))
#>               text sentiment prob_neg prob_neu prob_pos    class confidence
#> 1     I love this!      0.95     0.01     0.03     0.96 positive       0.96
#> 2        it's fine      0.10     0.08     0.74     0.18  neutral       0.74
#> 3 this is terrible     -0.93     0.95     0.03     0.02 negative       0.95
```

`sentiment` is `prob_pos - prob_neg`; `class` is the most probable class; `confidence` is
its probability. The probabilities are **calibrated** (temperature-scaled; expected
calibration error ≈ 0.015 on the held-out test), so a confidence means about what it says.

## `sentiment_match()`

The **same calibrated `sentiment`** as `sentiment_score()`, plus a nearest-phrase
explanation against **tunable poles**. The poles only shape the explanation
(`phrase` / `class` / `similarity`), never the score:

```r
my_categories <- list(positive = c("excited", "loving", "content", "happy"),
                      negative = c("lame", "lonely", "sad", "angry"))

sentiment_match(c("Will you marry me?", "Oh, you're breaking up with me..."),
                phrases = my_categories)
#>                                text sentiment phrase    class similarity
#> 1                Will you marry me?      0.54 loving positive       0.82
#> 2 Oh, you're breaking up with me...     -0.66    sad negative       0.79
```

Pass any set of poles, not just positive/negative, for arbitrary category matching, or
omit `phrases` to use the bundled balanced 40/40 default poles. Cosine similarity is
relative, so longer text matches any one phrase less strongly.

## `sentiment_provenance()`

See exactly what produced a score:

```r
sentiment_provenance("e5-small")
#> sentiment.ai provenance
#>   model    : e5-small (st, dim 384)
#>   prefix   : "query: "
#>   license  : MIT
#>   source   : https://huggingface.co/intfloat/multilingual-e5-small
#>   scoring  : mlp 1.0 (mlp, T=...)
```

# Calibration — how much to trust a score

`sentiment.ai` doesn't just pick a class — its probabilities are **calibrated**, so the
`confidence` returned by `sentiment()` means what it says: among the rows it scores at 0.8
confidence, about 80% really are that class. That makes `confidence` a number you can act
on — auto-accept the easy rows, send the unsure ones to a human.

On the held-out test set the **expected calibration error (ECE) is about 0.015** for
`e5-small` and **0.017** for `e5-base` — well-calibrated (an uncalibrated model is typically
0.05–0.15) — and it holds on real, non-synthetic text:

| stated confidence (e5-small) | actually correct |
|---:|---:|
| ~0.55 | 53% |
| ~0.70 | 73% |
| ~0.84 | 81% |
| ~0.90 | 92% |
| ~0.98 | 99% |

So you can triage on it:

```r
s      <- sentiment(reviews)
sure   <- s[s$confidence >= 0.85, ]   # trust these
review <- s[s$confidence <  0.85, ]   # route these to a human
```

Full method, per-class ECE, and Brier scores are in the reliability report. Calibration is
measured **in-domain** (reviews / short verbatims) on a single split, so treat very
different text — legal, code, heavy sarcasm — with more caution.

# Embeddings & matrix helpers

After `init_sentiment.ai()`, embed text yourself with `embed_text()` and compare with
`cosine()` / `cosine_match()`:

```r
target_mx <- embed_text(c("dogs", "cat", "IT", "computer"))
ref_mx    <- embed_text(c("animals", "technology"))
cosine_match(target_mx, ref_mx)[rank == 1]   # top match per row
```

# Python sibling

The same engine is on PyPI as `sentimentai-py`, shipping the **same** scoring heads
with a forward pass verified bit-for-bit against this package:

```bash
pip install --pre sentimentai-py
```

# Troubleshooting

This bridges R and Python via `reticulate`, so environment activation is the usual
snag. If reticulate won't switch environments (common in RStudio, projects, or
containers), it's normally because `RETICULATE_PYTHON` is set — restart R, set the
interpreter under *Tools > Global Options > Python*, or install into the base
`r-reticulate` environment. The legacy TensorFlow models additionally need a working
TensorFlow setup (`install_sentiment.ai(legacy = TRUE)`); the default e5 path avoids
all of that.
