---
title: "sentiment.ai"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sentiment.ai}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(sentiment.ai))
```

`sentiment.ai` turns text into sentiment scores using a sentence-embedding model
plus a small, bundled scoring head. The default backend is the on-device
multilingual model `multilingual-e5-small` — **no TensorFlow and no API key**.

**Project page:**
[https://benwiseman.github.io/sentiment.ai/](https://benwiseman.github.io/sentiment.ai/)

# Quick start

```r
# one-time setup of the (TensorFlow-free) Python backend
install_sentiment.ai()

# the model loads automatically on first use — no explicit init needed
sentiment_score(c("I love this!", "this is terrible"))
#> [1]  0.9  -0.9
```

Scores are rescaled to `[-1, 1]` — about `1` is positive, about `-1` is negative.
`init_sentiment.ai()` is optional; use it only for eager loading (e.g. at the top
of a benchmark script so the load time doesn't skew your first timing).

# Why embeddings?

Compared with lexicon/dictionary sentiment methods, this approach:

1. **Is more robust** — tolerates spelling mistakes and mixed case, and the default
   model handles ~100 languages.
2. **Doesn't need a rigid lexicon** — text is reduced to an embedding vector, so you
   get sensible scores for words and phrasings no dictionary lists.
3. **Lets you choose the context** — `sentiment_match()` lets you define what
   *positive* and *negative* mean for your domain.
4. **Is auditable** — scoring is deterministic, and `sentiment_provenance()` reports
   exactly which model and head produced a score.

# Installation & setup

After installing from CRAN, set up the Python backend once with
`install_sentiment.ai()`. The default install is TensorFlow-free. Useful arguments:

* `envname` — name of the Python environment (default `"r-sentiment-ai"`).
* `method` — `"auto"` (default), `"virtualenv"`, or `"conda"`.
* `gpu` — `FALSE` (default) installs the smaller **CPU-only** PyTorch; `TRUE` installs the
  CUDA build (Linux/Windows with an NVIDIA GPU). macOS is always CPU/MPS. The setup
  walkthrough offers this as a menu choice.
* `fast` — use `uv` (if installed) for a much faster install; falls back to pip safely.
* `legacy` — set `TRUE` to also install TensorFlow / TF-Hub for the opt-in legacy
  Universal Sentence Encoder models.

```r
install_sentiment.ai()                 # TensorFlow-free default
# install_sentiment.ai(legacy = TRUE)  # opt into the legacy USE models
```

Even simpler for newcomers: you can skip this step entirely. The first time you score
text in an interactive session, sentiment.ai offers a short setup walkthrough
(`setup_sentiment.ai()`) and — when [uv](https://docs.astral.sh/uv/) is installed — uses it
for a much faster install. In scripts / non-interactive sessions nothing is auto-installed;
you get a clear message to run `install_sentiment.ai()`.

Once the backend is set up you don't need to call `init_sentiment.ai()` yourself — the
model loads automatically on the first `sentiment_score()` / `sentiment()` /
`sentiment_match()` call. To load it eagerly the moment you attach the package (e.g. from
your `.Rprofile`), set `options(sentiment.ai.autoinit = TRUE)`.

# Choosing a model

`init_sentiment.ai(model = ...)` loads a model into memory:

* `"e5-small"` — default, 384-d, multilingual, on-device, no TensorFlow.
* `"e5-base"` — 768-d, best on-device quality.
* `"openai"` — `text-embedding-3-small` (paid API; text leaves your machine).
* `"en"` / `"en.large"` / `"multi"` / `"multi.large"` — legacy USE models, require
  `install_sentiment.ai(legacy = TRUE)`.

```r
init_sentiment.ai()                    # e5-small
init_sentiment.ai(model = "e5-base")   # larger on-device model
```

# Sentiment analysis

## `sentiment_score()`

The main function — one score per input. The scoring head is `"mlp"` (default) or
`"logistic"`; both are small pure-R JSON heads (no xgboost, no TensorFlow at score
time).

```r
sentiment_score(c("Will you marry me?", "Oh, you're breaking up with me..."))
```

## `sentiment()`

When you need the whole picture rather than one number — the per-class probabilities, a
label, and a confidence — use `sentiment()`. It returns a tidy data frame.

```r
sentiment(c("I love this!", "it's fine", "this is terrible"))
#>               text sentiment prob_neg prob_neu prob_pos    class confidence
#> 1     I love this!      0.95     0.01     0.03     0.96 positive       0.96
#> 2        it's fine      0.10     0.08     0.74     0.18  neutral       0.74
#> 3 this is terrible     -0.93     0.95     0.03     0.02 negative       0.95
```

`sentiment` is `prob_pos - prob_neg`; `class` is the most probable class; `confidence` is
its probability. The probabilities are **calibrated** (temperature-scaled; expected
calibration error ≈ 0.015 on the held-out test), so a confidence means about what it says.

## `sentiment_match()`

The **same calibrated `sentiment`** as `sentiment_score()`, plus a nearest-phrase
explanation against **tunable poles**. The poles only shape the explanation
(`phrase` / `class` / `similarity`), never the score:

```r
my_categories <- list(positive = c("excited", "loving", "content", "happy"),
                      negative = c("lame", "lonely", "sad", "angry"))

sentiment_match(c("Will you marry me?", "Oh, you're breaking up with me..."),
                phrases = my_categories)
#>                                text sentiment phrase    class similarity
#> 1                Will you marry me?      0.54 loving positive       0.82
#> 2 Oh, you're breaking up with me...     -0.66    sad negative       0.79
```

Pass any set of poles, not just positive/negative, for arbitrary category matching, or
omit `phrases` to use the bundled balanced 40/40 default poles. Cosine similarity is
relative, so longer text matches any one phrase less strongly.

## `sentiment_provenance()`

See exactly what produced a score:

```r
sentiment_provenance("e5-small")
#> sentiment.ai provenance
#>   model    : e5-small (st, dim 384)
#>   prefix   : "query: "
#>   revision : 614241f622f53c4eeff9890bdc4f31cfecc418b3
#>   license  : MIT
#>   source   : https://huggingface.co/intfloat/multilingual-e5-small
#>   scoring  : mlp 1.0 (mlp, T=1.281)
```

# Calibration — how much to trust a score

`sentiment.ai` doesn't just pick a class — its probabilities are **calibrated**, so the
`confidence` returned by `sentiment()` means what it says: among the rows it scores at 0.8
confidence, about 80% really are that class. That makes `confidence` a number you can act
on — auto-accept the easy rows, send the unsure ones to a human.

On the held-out test set the **expected calibration error (ECE) is about 0.015** for
`e5-small` and **0.017** for `e5-base` — well-calibrated (an uncalibrated model is typically
0.05–0.15) — and it holds on real, non-synthetic text:

| stated confidence (e5-small) | actually correct |
|---:|---:|
| ~0.55 | 53% |
| ~0.70 | 73% |
| ~0.84 | 81% |
| ~0.90 | 92% |
| ~0.98 | 99% |

So you can triage on it:

```r
s      <- sentiment(reviews)
sure   <- s[s$confidence >= 0.85, ]   # trust these
review <- s[s$confidence <  0.85, ]   # route these to a human
```

Full method, per-class ECE, and Brier scores are in the reliability report. Calibration is
measured **in-domain** (reviews / short verbatims) on a single split, so treat very
different text — legal, code, heavy sarcasm — with more caution.

# Picking a model

There are two on-device models. Both run with no TensorFlow, no API key, and handle
~100 languages.

| model | embedding dim | disk / RAM | relative speed | macro-F1 (real text) |
|:------|:---:|:---:|:---:|:---:|
| `e5-small` (**default**) | 384 | ~120 MB | baseline | 0.854 |
| `e5-base` | 768 | ~280 MB | ~2× slower | 0.899 |

**Use `e5-small` (the default) when:**

- You are scoring in a script or pipeline and throughput matters.
- You're on a laptop with limited RAM (< 8 GB).
- Your text is short — tweets, survey items, review snippets — where the smaller
  model loses little accuracy.
- You're exploring and want fast iteration.

**Upgrade to `e5-base` when:**

- You need the best possible accuracy and can afford the extra time/memory.
- Your text is longer or more nuanced (interview transcripts, open-ended verbatims,
  multi-sentence reviews).
- You are producing a final dataset for publication and the ~5 pp F1 gain matters.
- You have a GPU or are running on a server with ample RAM.

```r
# switch models explicitly
init_sentiment.ai(model = "e5-base")
sentiment_score("Surprisingly good for the price.")
#> [1] 0.72

# or set a persistent session default in your .Rprofile:
# options(sentiment.ai.model = "e5-base")
# The interactive setup wizard also lets you choose during install.
```

`e5-base` is not *always* better — on very short, simple text the gains are marginal,
and the slower throughput is a real cost. For most exploratory work `e5-small` is the
right choice; upgrade when you have a reason.

There is also an `openai` backend (`text-embedding-3-small` via the API) that matches
`e5-base` accuracy but sends text to OpenAI's servers and incurs per-token cost. Use it
only if you have an API key and are comfortable with data leaving your machine.

# Working with real data

The package ships `airline_tweets` — 14,640 tweets labelled negative / neutral /
positive. Here is a realistic end-to-end workflow.

```r
library(sentiment.ai)
data(airline_tweets)

# score all 14,640 tweets in one call (batched automatically)
scores <- sentiment(airline_tweets$text)
head(scores[, c("text", "sentiment", "class", "confidence")], 3)
#>                                                  text sentiment    class confidence
#> 1                @VirginAmerica What @dhepburn said.      0.07  neutral       0.71
#> 2 @VirginAmerica plus you've added commercials ...     -0.61 negative       0.83
#> 3   @VirginAmerica I didn't today... Must mean I ...    0.44 positive       0.72

# compare predicted class to the human label
scores$true_label <- airline_tweets$airline_sentiment
table(predicted = scores$class, true = scores$true_label)
#>           true
#> predicted  negative neutral positive
#>   negative     7673     322      260
#>   neutral       386     873      198
#>   positive      321     248     2359

# triage: auto-accept high-confidence rows, queue the rest for review
auto   <- scores[scores$confidence >= 0.85, ]
review <- scores[scores$confidence <  0.85, ]
cat(sprintf("%.0f%% auto-accepted, %.0f%% queued for review\n",
            100 * nrow(auto) / nrow(scores),
            100 * nrow(review) / nrow(scores)))
#> 71% auto-accepted, 29% queued for review
```

The `sentiment` column runs from −1 (negative) to +1 (positive); `confidence` is the
probability of the predicted class. Both are calibrated — see the Calibration section.

# Auditing your scores — `sentiment_diagnostics()`

`sentiment_diagnostics()` augments every row with five signals that tell you *when
not to trust the score*:

| signal | meaning |
|:---|:---|
| `entropy` | How spread is the probability mass? High = uncertain. |
| `confidence_band` | `"high"` / `"moderate"` / `"low"` calibrated from ECE data. |
| `mixed` | Both pos and neg mass > 0.25 — the text has competing signals. |
| `ood_similarity` | Cosine to the training-data centroid. Low = unlike training text. |
| `ood_flag` | `TRUE` when `ood_similarity < 0.20` — probably out-of-domain. |

```r
d <- sentiment_diagnostics(c(
  "I absolutely loved everything about this trip!",    # clear positive
  "Ein sehr ungewoehnlicher Text auf Deutsch.",        # German (in-domain: e5 is multilingual)
  "0x4A 0x6F 0x65",                                   # hex bytes: out-of-domain
  "The food was great but the service was terrible."   # mixed signal
))
d[, c("text", "sentiment", "confidence_band", "mixed", "ood_flag")]
#>                                      text sentiment confidence_band mixed ood_flag
#> 1    I absolutely loved everything ...      0.97            high FALSE    FALSE
#> 2    Ein sehr ungewoehnlicher Text ...      0.05        moderate FALSE    FALSE
#> 3                         0x4A 0x6F ...     -0.12             low FALSE     TRUE
#> 4    The food was great but the se...      0.12        moderate  TRUE    FALSE
```

The recommended **auto-accept triage rule**:

```r
auto_accept <- d$confidence_band >= "moderate" & !d$mixed & !is.na(d$ood_flag) & !d$ood_flag
```

`confidence_band` is an ordered factor (`low < moderate < high`), so `>=` comparisons work
naturally. Route rows where `auto_accept == FALSE` to a human reviewer.

# Validating against human labels — `sentiment_agreement()`

When you need a number you can quote in a methods section, use `sentiment_agreement()`.
It compares the model's scores to a set of human-provided reference labels and returns
the standard agreement statistics used in NLP and psychometrics.

```r
# Using the airline_tweets dataset: compare to the crowd-sourced labels
scores <- sentiment_score(airline_tweets$text)
labels <- airline_tweets$airline_sentiment   # "negative" / "neutral" / "positive"

ag <- sentiment_agreement(scores, labels)
print(ag)
#> sentiment.ai agreement statistics  (n = 14640)
#> -----------------------------------------------
#>   Spearman r (score vs label)  : 0.721
#>   Percent agreement (3-class)  : 74.6%
#>   Weighted kappa (quad, 3-cls) : 0.682
#>
#> Human-human ceiling (indicative): weighted kappa ~0.50-0.65 on
#> GoEmotions / SemEval-2017-4.
#>   Krippendorff alpha (ordinal) : 0.612
#>   ICC(2,1)                     : 0.718   95% CI [0.710, 0.726]
#>
#> Confusion matrix (rows=true, cols=predicted):
#>           predicted
#> true       negative neutral positive
#>   negative     7673     386      321
#>   neutral       322     873      248
#>   positive      260     198     2359
```

The **weighted kappa of 0.68** exceeds the typical human-human ceiling for 3-class
sentiment (0.50–0.65), which is the number to quote in a methods section. You can
also adjust the neutral-zone boundaries:

```r
# tighter neutral zone (scores in (-0.2, 0.2) become "neutral")
ag2 <- sentiment_agreement(scores, labels,
                           positive_threshold =  0.2,
                           negative_threshold = -0.2)
```

# Embeddings & matrix helpers

After `init_sentiment.ai()`, embed text yourself with `embed_text()` and compare with
`cosine()` / `cosine_match()`:

```r
target_mx <- embed_text(c("dogs", "cat", "IT", "computer"))
ref_mx    <- embed_text(c("animals", "technology"))
cosine_match(target_mx, ref_mx)[rank == 1]   # top match per row
#>      target  reference similarity rank
#> 1:     dogs    animals       0.72    1
#> 2:      cat    animals       0.68    1
#> 3:       IT technology       0.65    1
#> 4: computer technology       0.61    1
```

For large reference sets (thousands of phrases), add `approx = TRUE` for a much faster
approximate nearest-neighbour search via `RANN`:

```r
cosine_match(target_mx, large_ref_mx, approx = TRUE)
```

# Python sibling

The same engine is on PyPI as `sentimentai-py`, shipping the **same** scoring heads
with a forward pass verified bit-for-bit against this package:

```bash
pip install --pre sentimentai-py
```

# Troubleshooting

This bridges R and Python via `reticulate`, so environment activation is the usual
snag.

**Most common fix — `RETICULATE_PYTHON` is set:**

```r
# 1. Check whether the variable is pointing at the wrong interpreter:
Sys.getenv("RETICULATE_PYTHON")

# 2. If it is, clear it and restart R, then re-run install_sentiment.ai():
Sys.unsetenv("RETICULATE_PYTHON")
# then: restart R (Session > Restart R in RStudio) and run install_sentiment.ai()
```

In RStudio, the most reliable fix is *Tools > Global Options > Python > Select
Interpreter* — point it at the `r-sentiment-ai` virtualenv. This writes the path
to your user profile so it survives restarts.

**Other common causes:**

- *Wrong working directory or an `.Rprofile`/`.Renviron` in the project root* setting
  `RETICULATE_PYTHON` — check with `readLines(".Renviron")`.
- *Running inside a Docker container or CI* with a system Python active — pass
  `envname = "r-sentiment-ai"` explicitly to `init_sentiment.ai()`.
- *Legacy TensorFlow models* need `install_sentiment.ai(legacy = TRUE)`; the default
  e5 path avoids TensorFlow entirely.
