---
image: "media/nnet_big.svg"
title: sentiment.ai
subtitle: Sentiment analysis from sentence embeddings — multilingual, on-device, TensorFlow-free
output:
  html_document:
    toc: yes
    toc_float:
        collapsed: true
        smooth_scroll: true
    theme: flatly
    code_folding: show
    highlight: espresso
    css: www/css/custom.css
    includes:
      in_header: header.html
  word_document:
    toc: yes
editor_options:
  chunk_output_type: console
always_allow_html: true
---

<script>
  $(document).ready(function() {
    $('#TOC').parent().prepend('<div id=\"nav_logo\"><img src=\"media/sentimentai.png"></div>');
  });
</script>

```{css, echo=FALSE}
 #TOC::before {
  font-size: 32px;
  font-weight: 900;
  text-align: center;
  content: "sentiment.ai";
  display: block;
  width: 200px;
  height: 80px;
  line-height: 80px;
  margin: 10px 10px 10px 20px;
  background-size: contain;
  background-position: center center;
  background-repeat: no-repeat;
}
#TOC{
    border:none;
}
.new-in-v2 {
  background: #f0f7ff;
  border-left: 4px solid #2c7bb6;
  padding: 12px 16px;
  border-radius: 4px;
  margin: 16px 0;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- [START BADGES] -->
<p align="left">
  <a href="https://CRAN.R-project.org/package=sentiment.ai"><img src="https://img.shields.io/cran/v/sentiment.ai?style=flat-square" alt="CRAN version" /></a>
  <a href="https://CRAN.R-project.org/package=sentiment.ai"><img src="https://img.shields.io/cran/l/sentiment.ai?style=flat-square" alt="MIT License" /></a>
  <a hidden href="https://hits.seeyoufarm.com"><img src="https://hits.seeyoufarm.com/api/count/incr/badge.svg?url=https%3A%2F%2Fgithub.com%2FBenWiseman%2Fsentiment.ai&count_bg=%23AB6EE5&title_bg=%23555555&icon=&icon_color=%23E7E7E7&title=hits&edge_flat=false"/></a>
</p>
<!-- [END BADGES] -->

# Overview

`sentiment.ai` turns text into sentiment scores using a sentence-embedding model plus
a small, bundled scoring head. The default backend is the on-device multilingual model
**`multilingual-e5-small`** — **no TensorFlow, no API key, no data leaving your machine**.

<div class="new-in-v2">
**What's new in v2 (1.0.1)**

- **No TensorFlow** — default install is ~120 MB on-device, Apple-Silicon-clean.
- **Calibrated confidence** — ECE ≈ 0.015; `confidence` actually means what it says.
- **`sentiment()`** — tidy 3-class output (text, score, prob_neg/neu/pos, class, confidence).
- **`sentiment_diagnostics()`** — know *when not to trust* a score: entropy, mixed-signal flag, out-of-domain detector.
- **`sentiment_agreement()`** — validate against human labels; get weighted-κ, Krippendorff's α, ICC.
- **Automatic dependency fallback** — if a "latest" pip install breaks, `install_sentiment.ai()` snaps to the verified baseline automatically.
- **Model selection in the setup wizard** — choose e5-small vs e5-base at install time.
</div>

Compared with lexicon/dictionary methods:

1. **More robust** — handles spelling mistakes, mixed case, and ~100 languages from one model.
2. **No rigid lexicon** — text becomes an embedding vector, so novel phrasing and domain jargon are handled naturally.
3. **Tunable context** — `sentiment_match()` lets you define what *positive* and *negative* mean for your domain.
4. **Auditable** — scoring is deterministic; `sentiment_provenance()` reports the exact model, revision SHA, and scoring head behind every score.

---

# Getting started {.tabset .tabset-fade .tabset-pills}

## R users

```{r r-quickstart, echo=TRUE, eval=FALSE}
install.packages("sentiment.ai")        # from CRAN
library(sentiment.ai)

# one-time setup — walks you through it interactively
install_sentiment.ai()

# the model loads on first use — no explicit init() needed
sentiment_score(c("I love this!", "this is terrible"))
#> [1]  0.9  -0.9

# full 3-class tidy output
sentiment(c("I love this!", "it's fine", "this is terrible"))
#>               text sentiment    class confidence
#> 1     I love this!      0.95 positive       0.96
#> 2        it's fine      0.10  neutral       0.74
#> 3 this is terrible     -0.93 negative       0.95
```

`init_sentiment.ai()` is optional — only needed for eager loading (e.g. to avoid a
slow first call inside a benchmark loop). The model loads automatically on the first
`sentiment_score()` / `sentiment()` / `sentiment_match()` call.

**Persistent model default:**
```{r r-default-model, echo=TRUE, eval=FALSE}
# In your .Rprofile — picks up at every library(sentiment.ai)
options(sentiment.ai.model = "e5-base")
```

## Python users

```bash
pip install --pre sentimentai-py
```

```python
import sentimentai as sa

# simple score
sa.sentiment_score(["I love this!", "this is terrible"])
# [-0.9, 0.9]  -- note: Python returns [neg, pos] ordering

# tidy output
sa.sentiment(["I love this!", "it's fine", "this is terrible"])
#    text                  sentiment  class     confidence
# 0  I love this!          0.95       positive  0.96
# 1  it's fine             0.10       neutral   0.74
# 2  this is terrible     -0.93       negative  0.95

# same scoring heads as the R package — results are bit-for-bit identical
sa.sentiment_provenance("e5-small")
```

Install options:
```bash
pip install --pre sentimentai-py                  # CPU (default)
pip install --pre sentimentai-py[openai]          # + OpenAI API support
```

---

# A quick taste

Some real scores from the default model — it reads context, not just keywords:

| text | `sentiment.ai` |
|:-----|:--------------:|
| the resturant is my favorite! | **+0.96** |
| this restront is my FAVRIT innit! | **+0.46** |
| I love watching scary horror movies | **+0.88** |
| I had **a blast** on my trip to Nagasaki | **+0.24** |
| **The blast** in Nagasaki | **−0.99** |
| my absolute favorite until they gave me food poisoning | **−0.96** |
| What a great car. It stopped working after a week. | **−0.52** |

(Default `e5-small`. Note "**a blast** on my trip" vs "**The blast** in Nagasaki".
Scoring is deterministic; `sentiment_provenance()` logs exactly what produced each score.)

---

# Why not just use tidytext / vader / sentimentr?

Lexicon-based tools score words against a fixed dictionary. They are fast and easy to
inspect, but share hard limits: out-of-vocabulary terms are missed, negation is handled
only by hand-written rules, and most convenient lexicons are English-only.
`sentiment.ai` maps the whole sentence to an embedding vector and classifies that vector:

- **Handles ~100 languages** with one model, no separate multilingual lexicons.
- **Scores phrases no dictionary has ever seen** — slang, domain jargon, product names.
- **Explicit neutral class** — not just "not positive, not negative".
- **Calibrated confidence** (ECE ≈ 0.015) so you can triage on it.
- **Tunable context** — redefine positive/negative for your domain.

The tradeoff is setup time: lexicon tools install with no Python; `sentiment.ai` needs
a one-time ~120 MB model download. After that there is no internet connection required
and scoring is deterministic across machines.

---

# Picking a model {.tabset .tabset-fade .tabset-pills}

## e5-small (default)

| property | value |
|:---------|:------|
| Embedding dim | 384 |
| Disk / RAM | ~120 MB |
| Languages | ~100 |
| Speed | ~310 texts/sec (CPU) |
| Macro-F1 — shipped MLP head | 0.734 (mixed-domain held-out, n=1,247) |
| Macro-F1 — training benchmark | 0.854 (XGBoost scorer; see note) |

**Best for:** scripts and pipelines where throughput matters, laptops with < 8 GB RAM,
short text (tweets, survey items, review snippets), exploratory work and fast iteration.

```{r e5small, echo=TRUE, eval=FALSE}
sentiment_score(texts)                  # uses e5-small by default
sentiment_score(texts, model="e5-small")  # explicit
```

## e5-base (higher accuracy)

| property | value |
|:---------|:------|
| Embedding dim | 768 |
| Disk / RAM | ~280 MB |
| Languages | ~100 |
| Speed | ~141 texts/sec (CPU) |
| Macro-F1 — shipped MLP head | 0.823 (mixed-domain held-out, n=1,247) |
| Macro-F1 — training benchmark | 0.899 (XGBoost scorer; see note) |

**Best for:** final datasets for publication, longer/more nuanced text (interview transcripts,
open-ended verbatims), servers with ample RAM, cases where the ~9 pp F1 gain over e5-small matters.

```{r e5base, echo=TRUE, eval=FALSE}
sentiment_score(texts, model = "e5-base")

# make e5-base your session default
init_sentiment.ai(model = "e5-base")

# or permanent (in .Rprofile):
options(sentiment.ai.model = "e5-base")
```

## OpenAI API

| property | value |
|:---------|:------|
| Embedding dim | 1536 |
| Model | `text-embedding-3-small` |
| Speed | network-dependent |
| Macro-F1 — training benchmark | 0.886 (XGBoost scorer; see note) |
| Cost | per-token API charge |

Text leaves your machine and is sent to OpenAI's servers. Requires an API key.
On the training benchmark (XGBoost scorer), OpenAI is marginally behind `e5-base` (0.886
vs 0.899) — suggesting the e5 embedding space suits the scoring head better.

```{r openai, echo=TRUE, eval=FALSE}
Sys.setenv(OPENAI_API_KEY = "sk-...")
sentiment_score(texts, model = "openai")
```

---

# Sentiment analysis — R {.tabset .tabset-fade .tabset-pills}

## `sentiment_score()`

One score per input in `[-1, 1]`.

```{r score, echo=TRUE, eval=FALSE}
sentiment_score(c("Will you marry me?", "Oh, you're breaking up with me..."))
#> [1]  0.54 -0.66
```

## `sentiment()`

Tidy data frame with the full 3-class signal. Use when the neutral mass matters or to
triage low-confidence rows.

```{r tidy, echo=TRUE, eval=FALSE}
sentiment(c("I love this!", "it's fine", "this is terrible"))
#>               text sentiment prob_neg prob_neu prob_pos    class confidence
#> 1     I love this!      0.95     0.01     0.03     0.96 positive       0.96
#> 2        it's fine      0.10     0.08     0.74     0.18  neutral       0.74
#> 3 this is terrible     -0.93     0.95     0.03     0.02 negative       0.95

# triage: trust high-confidence rows automatically
s    <- sentiment(my_reviews)
sure <- s[s$confidence >= 0.85, ]   # auto-accept
flag <- s[s$confidence <  0.85, ]   # route to human
```

## `sentiment_match()`

Same calibrated score as `sentiment_score()`, plus a nearest-phrase explanation against
**tunable poles**. The poles define what *positive* and *negative* mean for your domain.

```{r match, echo=TRUE, eval=FALSE}
sentiment_match(c("Will you marry me?", "Oh, you're breaking up with me..."),
                phrases = list(
                  positive = c("excited", "loving", "content", "happy"),
                  negative = c("lame",    "lonely", "sad",     "angry")))
#>                                text sentiment phrase    class similarity
#> 1                Will you marry me?      0.54 loving positive       0.82
#> 2 Oh, you're breaking up with me...     -0.66    sad negative       0.79
```

Omit `phrases` to use the bundled balanced 40/40 default poles.

## `sentiment_provenance()`

See exactly what produced a score.

```{r prov, echo=TRUE, eval=FALSE}
sentiment_provenance("e5-small")
#> sentiment.ai provenance
#>   model    : e5-small (st, dim 384)
#>   prefix   : "query: "
#>   revision : 614241f622f53c4eeff9890bdc4f31cfecc418b3
#>   license  : MIT
#>   source   : https://huggingface.co/intfloat/multilingual-e5-small
#>   scoring  : mlp 1.0 (mlp, T=1.281)
```

---

# Sentiment analysis — Python {.tabset .tabset-fade .tabset-pills}

## `sentiment_score()`

```python
import sentimentai as sa
scores = sa.sentiment_score(["I love this!", "this is terrible"])
# [0.95, -0.93]
```

## `sentiment()`

```python
df = sa.sentiment(["I love this!", "it's fine", "this is terrible"])
print(df[["text", "sentiment", "class", "confidence"]])
#               text  sentiment     class  confidence
# 0     I love this!       0.95  positive        0.96
# 1        it's fine       0.10   neutral        0.74
# 2 this is terrible      -0.93  negative        0.95
```

## `sentiment_match()`

```python
poles = {
    "positive": ["excited", "loving", "content", "happy"],
    "negative": ["lame",    "lonely", "sad",     "angry"]
}
df = sa.sentiment_match(["Will you marry me?", "Breaking up with me..."],
                        phrases=poles)
```

## `sentiment_provenance()`

```python
sa.sentiment_provenance("e5-small")
# sentiment.ai provenance
#   model    : e5-small (st, dim 384)
#   revision : 614241f622f53c4eeff9890bdc4f31cfecc418b3
#   scoring  : mlp 1.0
```

---

# Calibration — how much to trust a score

`sentiment.ai` doesn't just pick a class — its probabilities are **calibrated**, so the
`confidence` from `sentiment()` means what it says: among rows scored at 0.8 confidence,
about 80% really are that class.

On the held-out test set the **expected calibration error (ECE) is about 0.015**
(`e5-small`) / **0.017** (`e5-base`) — well-calibrated (an uncalibrated model is typically
0.05–0.15):

| stated confidence (e5-small) | actually correct |
|---:|---:|
| ~0.55 | 53% |
| ~0.70 | 73% |
| ~0.84 | 81% |
| ~0.90 | 92% |
| ~0.98 | 99% |

Calibration was measured in-domain (reviews / short verbatims, single split). Treat very
different text — legal documents, code, heavy sarcasm — with more caution.

---

# Diagnostics — when not to trust a score

`sentiment_diagnostics()` augments every row with signals that say *when not to trust it*:

| column | meaning |
|:-------|:--------|
| `entropy` | Shannon entropy of the 3-class probs (nats). High = head is uncertain. |
| `confidence_band` | Ordered factor `"low"` / `"moderate"` / `"high"` calibrated from ECE data. |
| `mixed` | `TRUE` when both `prob_pos` and `prob_neg` > 0.25 — competing signals. |
| `ood_similarity` | Cosine to training centroids. Low (~< 0.20) = unlike training data. |
| `ood_flag` | `TRUE` when `ood_similarity < 0.20` — probably out-of-domain. |

```{r diag-r, echo=TRUE, eval=FALSE}
d <- sentiment_diagnostics(c(
  "I absolutely loved everything about this trip!",
  "Ein sehr ungewoehnlicher Text auf Deutsch.",      # German — in-domain (e5 is multilingual)
  "0x4A 0x6F 0x65",                                 # hex bytes — out-of-domain
  "The food was great but the service was terrible." # mixed signal
))
d[, c("text", "sentiment", "confidence_band", "mixed", "ood_flag")]
```

**Auto-accept triage rule:**
```{r triage, echo=TRUE, eval=FALSE}
auto_accept <- d$confidence_band >= "moderate" & !d$mixed &
               !is.na(d$ood_flag) & !d$ood_flag
# TRUE rows: route to auto-pipeline
# FALSE rows: send to human review
```

`confidence_band` is an ordered factor (`low < moderate < high`), so `>=` comparisons work naturally.

---

# Validating against human labels

`sentiment_agreement()` compares model scores to human-provided labels and returns the
statistics needed for a methods section.

```{r agree-r, echo=TRUE, eval=FALSE}
data(airline_tweets)
scores <- sentiment_score(airline_tweets$text)
ag     <- sentiment_agreement(scores, airline_tweets$airline_sentiment)
print(ag)
#> sentiment.ai agreement statistics  (n = 14640)
#> -----------------------------------------------
#>   Spearman r (score vs label)  : 0.721
#>   Percent agreement (3-class)  : 74.6%
#>   Weighted kappa (quad, 3-cls) : 0.682
#>
#> Human-human ceiling (indicative): weighted kappa ~0.50-0.65 on
#> GoEmotions / SemEval-2017-4.
#>   Krippendorff alpha (ordinal) : 0.612
#>   ICC(2,1)                     : 0.718   95% CI [0.710, 0.726]
```

A **weighted kappa of 0.68** exceeds the typical human-human ceiling on 3-class sentiment
(0.50–0.65) — this is the number to quote in a methods section.

Adjust the neutral-zone boundaries to match your application:
```{r agree-thresh, echo=TRUE, eval=FALSE}
ag2 <- sentiment_agreement(scores, labels,
                           positive_threshold =  0.2,
                           negative_threshold = -0.2)
```

---

# All modes at a glance

| mode | `model=` | requires | embed dim | model download | storage / 1M texts |
|:-----|:---------|:---------|----------:|--------------:|-------------------:|
| **on-device default** | `"e5-small"` | PyTorch CPU | 384 | ~120 MB | ~1.5 GB |
| **on-device quality** | `"e5-base"` | PyTorch CPU | 768 | ~280 MB | ~3.0 GB |
| **API** | `"openai"` | API key + internet | 1536 | none | ~6.0 GB |
| **legacy** | `"en"` / `"en.large"` / `"multi"` | TensorFlow | 512 | varies | ~2.0 GB |

The `mlp`/`logistic` scoring heads ship inside the package for the two on-device modes —
no additional download. Storage figures are raw float32; `saveRDS(..., compress="xz")` cuts
this roughly 3–4× in practice.

---

# Benchmarks {.tabset .tabset-fade .tabset-pills}

**Setup:** 1,000 airline tweets (seed 42); macOS aarch64 M-series, 8 CPU cores, no GPU.
Accuracy = macro-F1 (3-class: neg/neu/pos) against crowd labels. Speed = median of 3 runs.

Three accuracy sources — each labelled clearly:

- **†** = 1,000 airline tweet sample (cross-domain; models trained on Amazon/IMDB/financial, *not* Twitter)
- **★** = same mixed-domain held-out set (n = 1,247 real-only; shipped MLP head; bootstrap CIs available)
- **☆** = training benchmark (XGBoost scorer on full training data — stronger scorer than shipped; for reference)

## R — speed

| method | texts/sec | type |
|:-------|----------:|:-----|
| tidytext (bing) | **32,258** | pure R, dictionary lookup |
| syuzhet (AFINN) | 3,311 | pure R, word-score sum |
| sentimentr | 564 | R + C++, sentence-aware |
| **sentiment.ai e5-small** | **310** | neural, CPU, MLP head |
| **sentiment.ai e5-base** | **141** | neural, CPU, 768-d |
| sentiment.ai openai | ~300–600 t/s | API, rate-limited |

## R — accuracy

| method | macro-F1 (†) | macro-F1 (★) | macro-F1 (☆) |
|:-------|:-----------:|:------------:|:------------:|
| **sentiment.ai e5-base** | 0.595 | **0.823** | 0.899 |
| sentiment.ai e5-small | 0.613 | **0.734** | 0.854 |
| sentiment.ai openai | — | — | 0.886 |
| twitter-roberta | 0.777 ⚠️ | **0.656** | — |
| tidytext (bing) | 0.511 | — | — |
| syuzhet (AFINN) | 0.301 | — | — |
| sentimentr | 0.293 | — | — |

⚠️ roberta airline-tweet figure is near-in-domain (TweetEval training overlap). Its ★ result
(0.656) is the fair out-of-sample number.

**Bootstrap CIs on ★ (2,000 resamples):**

- e5-base vs roberta: Δ = **+0.167**, 95% CI [+0.134, +0.199] — entirely above zero
- e5-small vs roberta: Δ = **+0.078**, 95% CI [+0.048, +0.107] — entirely above zero
- e5-base vs e5-small: Δ = **+0.089**, 95% CI [+0.061, +0.118]

**Note on ★ vs ☆ numbers:** ★ measures the *shipped MLP head* on the held-out set. ☆ measures
the *XGBoost scorer* trained on all available data — a stronger scorer that is no longer the
default. The gap (0.823 vs 0.899 for e5-base) reflects the XGBoost head being stronger, not a
change in embedding quality. Even with the shipped MLP head, both models beat roberta with
statistical significance.

## Python — speed

| method | texts/sec | type |
|:-------|----------:|:-----|
| vaderSentiment | **4,875** | pure Python, rule-based |
| TextBlob | 4,182 | NaiveBayes + pattern |
| **sentimentai-py e5-small** | **523** | neural, sentence-transformers |
| **sentimentai-py e5-base** | **177** | neural, 768-d |
| HF twitter-roberta | 19 | fine-tuned transformer, CPU |

`sentimentai-py` is slightly faster than the R package (~523 vs ~310 t/s for e5-small)
because it avoids the R↔Python round-trip.

## Python — accuracy

| method | macro-F1 (†) | notes |
|:-------|:-----------:|:------|
| HF twitter-roberta | 0.777 ⚠️ | near-in-domain on airline tweets |
| **sentimentai-py e5-small** | **0.615** | cross-domain transfer |
| **sentimentai-py e5-base** | **0.596** | cross-domain transfer |
| vaderSentiment | 0.473 | airline vocab mismatch |
| TextBlob | 0.349 | general-purpose |

⚠️ roberta airline-tweet result is near-in-domain. Its ★ result (0.656) is the fair number.
On the mixed-domain held-out set, **both sentiment.ai models beat roberta with statistical
significance** (CIs entirely above zero — see R accuracy tab for the full bootstrap results).

## Multilingual

Corpus: 892 GPT-4o-mini synthetic examples (50 per class per language); macOS aarch64 CPU.
*vader and roberta are English-only — non-English scores are shown to illustrate their limits.*

| method | overall | english | spanish | french | german | portuguese | arabic |
|:-------|:-------:|:-------:|:-------:|:------:|:------:|:----------:|:------:|
| **sentimentai e5-base** | **0.827** | **0.890** | **0.926** | **0.868** | **0.870** | 0.560 | **0.791** |
| sentimentai e5-small | 0.761 | 0.865 | 0.787 | 0.784 | 0.747 | 0.570 | 0.766 |
| vader | 0.432 | 0.710 | 0.484 | 0.322 | 0.291 | 0.271 | 0.457 |
| twitter-roberta | N/A | 0.837 | *N/A* | *N/A* | *N/A* | *N/A* | *N/A* |

On English, e5-base (0.890) edges roberta (0.837). On every other language, roberta and vader
cannot compete. vader collapses to F1=0.271 on German (roughly chance-level on 3-class).

**Portuguese note:** both e5 models score ~0.56 on Portuguese — a genuine gap worth
acknowledging. Possible causes: Portuguese underrepresentation in the e5 training mix, or
synthetic corpus artefacts. Use `e5-base` for Portuguese but validate on your own data.

*Corpus note:* synthetic text generated by GPT-4o-mini tends to be cleaner and less ambiguous
than real-world text. Absolute F1 figures are optimistic vs messy real text; the relative
ordering across languages and methods is the reliable signal.

## When to use each

| | e5-small | e5-base | twitter-roberta | vaderSentiment |
|:--|:--:|:--:|:--:|:--:|
| Cross-domain (★ F1) | 0.734 | **0.823** | 0.656 | — |
| Multilingual (6 lang) | **✓** | **✓** | ✗ English only | ✗ English only |
| Calibrated confidence | **✓** | **✓** | ✗ | ✗ |
| Tunable poles | **✓** | **✓** | ✗ | ✗ |
| Speed (CPU) | ~500 t/s | ~175 t/s | ~20 t/s | ~5,000 t/s |
| No download | ✗ | ✗ | ✗ (~500 MB) | **✓** |
| No API cost | **✓** | **✓** | **✓** | **✓** |

---

# Installation & setup {.tabset .tabset-fade .tabset-pills}

## R

```{r r-install, echo=TRUE, eval=FALSE}
install.packages("sentiment.ai")   # from CRAN
library(sentiment.ai)

# one-time setup — the interactive wizard walks you through it,
# including choosing e5-small vs e5-base as your default model
install_sentiment.ai()
```

**Key arguments:**

| argument | default | effect |
|:---------|:-------:|:-------|
| `method` | `"auto"` | `"virtualenv"` or `"conda"` |
| `gpu` | `FALSE` | `TRUE` = CUDA build (Linux/Windows with NVIDIA GPU) |
| `pin_versions` | `NA` | `NA` = try latest, auto-fallback to verified baseline on failure; `TRUE` = always use pinned; `FALSE` = skip check |
| `legacy` | `FALSE` | `TRUE` = also install TensorFlow / USE legacy stack |
| `fast` | `NA` | uses `uv` if on PATH for a much faster install |

**Troubleshooting (most common fix):**
```{r r-troubleshoot, echo=TRUE, eval=FALSE}
# Check if RETICULATE_PYTHON is pointing at the wrong interpreter
Sys.getenv("RETICULATE_PYTHON")

# If set, clear it and restart R
Sys.unsetenv("RETICULATE_PYTHON")
# then restart R and re-run install_sentiment.ai()
```

In RStudio: *Tools > Global Options > Python > Select* — point it at the
`r-sentiment-ai` virtualenv.

## Python

```bash
pip install --pre sentimentai-py
```

```python
import sentimentai as sa

# first use triggers a ~120 MB model download (one-time)
sa.sentiment_score(["Hello world"])

# to use the larger, more accurate model:
sa.init(model="e5-base")
sa.sentiment_score(["Hello world"], model="e5-base")
```

Requirements: Python 3.8+, PyTorch (CPU or CUDA).

---

# Embeddings & matrix helpers

Embed text yourself with `embed_text()` and compare with `cosine()` / `cosine_match()`:

```{r matrix, echo=TRUE, eval=FALSE}
target_mx <- embed_text(c("dogs", "cat", "keyboard", "mouse"))
ref_mx    <- embed_text(c("animals", "technology"))

cosine_match(target_mx, ref_mx)[rank == 1]
#>     target  reference similarity rank
#> 1:    dogs    animals       0.72    1
#> 2:     cat    animals       0.68    1
#> 3: keyboard technology       0.65    1
#> 4:   mouse technology       0.57    1

# approx=TRUE for large reference sets (requires RANN; much faster)
cosine_match(target_mx, large_ref_mx, approx = TRUE)
```

---

# Contribute a scoring head

Heads live under `inst/scoring/<type>/<version>/<model>.json`. A head is a small JSON
describing the forward pass (`{type, dim, T, layers | coef, classes}`), evaluated by a
pure-R scorer — no xgboost or TensorFlow at score time.

Use `head_path =` on `sentiment_score()` / `sentiment()` / `sentiment_diagnostics()` to
load a custom or domain-retrained head without touching the installed package:

```{r custom-head, echo=TRUE, eval=FALSE}
# point directly at a custom JSON head
sentiment_score(texts, head_path = "~/my_domain_head.json")
sentiment_diagnostics(texts, head_path = "~/my_domain_head.json")
```

---

<br><hr><br>

Originally created by the [Korn Ferry Institute](https://www.kornferry.com/institute) AITMI team. MIT licensed.
